Lazy Linearity for a Core Functional Language

### 2025-10-17 popl26.tex
.

\item We introduce Linear Core, a \ROUNDTWO{lazy} linear language with
the key features of Core (except for type equality coercions), whose type
system \ROUNDTWO{accepts the lazy} semantics of linearity. To the best of our
knowledge, this is the first type system \ROUNDTWO{to accept programs
which depend on non-strict evaluation to be understood as linear}
(Section~\ref{sec:main:linear-core}).

@@

Additional proofs and definitions are given in \ROUNDTWO{TODO}
Appendices~\ref{app:proofs},~\ref{app:optimisations} and~\ref{app:opsem}.

@@

Linear type systems guarantee that linear resources are consumed exactly once.
\ROUNDTWO{Thus}, whether a program is \ROUNDTWO{accepted as linear} intrinsically depends on
the definition of consuming a resource.
%
\ROUNDTWO{Traditionally}, consuming a resource \ROUNDTWO{has been} equated with its 
syntactic occurrence along a control-flow path in the program. However, as this
section \ROUNDTWO{aims to} make clear, \ROUNDTWO{in a non-strict context, a
syntactic definition of linearity is much too conservative}.
%
For instance, in the following Haskell-like program, \lstinline{f} is a linear
function (a function that must use its argument
exactly once), but there are two syntactic
occurrences of \lstinline{handle}: one is in a let-bound computation that
closes this handle, the other \ROUNDTWO{is in the first branch}. If the program is
evaluated lazily, the handle will, \ROUNDTWO{dynamically, either be closed in
the second branch, or be returned untouched in the first}, since the computation
which closes the handle is \ROUNDTWO{not needed in the first branch}.
%
\begin{lstlisting}
f : Handle *' $\lolli$ '* Handle
f handle =
    let closed = close handle in
    if <cond>
        then handle
        else closed
\end{lstlisting}
%
% Intuitively, a computation that depends on a linear resource to produce a
% result consumes that resource iff the result is effectively computed; in
% contrast, a computation that depends on a linear resource, but is never run,
% will not consume that resource.
%
% Even though there are two occurrences of \lstinline{handle} in the program, and
% it looks as though the \lstinline{handle} may be closed before returned, this function may
% actually only \emph{consume} used once -- it depends on the evaluation strategy of the
% language -- the \lstinline{handle} is closed iff the let bound computation is
% evaluated.
%
%
% Intuitively, \emph{consuming} a resource is not necessarily synonymous with its
% syntactic occurrence. In the example, \emph{non-strict} evaluation means the
% handle is never closed by the function.
%
%
% Depending on the evaluation strategy of the language, a computation that closes
% a handle might not be evaluated, and thus is not consumed.
% With \emph{eager} evaluation semantics,
% the let
%   bound expression \lstinline{close handle} is eagerly evaluated, and the \lstinline{handle} will be closed before being
% returned. It is clear that a linear type system should not accept such a
% program since the linear resource \lstinline{handle} is duplicated -- it is used in a
% computation that closes it, while still being made available to the caller of
% the function;
% %
% on the other hand, with \emph{call-by-need},
%
% Reconsider the \lstinline{handle} example program under
% \emph{call-by-need} evaluation semantics:
% The let bound expression is only evaluated if the binding \lstinline{x} is
% needed. The function returns the \lstinline{handle} right away, and forgets the
% let bound \lstinline{x} as its scope is exited, such that \lstinline{x} will
% never be needed -- consequently, the \lstinline{handle} is never closed by the
% function...
%
%
%

@@

\subsubsection{Branching on WHNF-ness}

@@

\subsubsection{Proof irrelevant resources}

>>>>

\item We introduce Linear Core, a \ROUNDTWO{lazy} linear language with
the key features of Core (except for type equality coercions), whose type
system \ROUNDTWO{is designed to reconcile with the lazy} semantics of
linearity. \ROUNDTWO{Our system achieves this via a novel combination
  of features: explicit tracking linear dependencies of variables
  through \emph{usage environments}
  (Section~\ref{sec:usage-environments}), codifying that lazy variable
  bindings do not consume resources upon definition, but rather when
  the variables are consumed;
  a distinct treatment of case
  scrutinees in weak-head normal form (Section~\ref{sec:onwhnf}),
  identifying that linear resources captured by case expressions in
  WHNF can be safely used in branches since the case will not further
  evaluate its scrutinee; and a notion of irrelevant and tagged resources
  (Section~\ref{sec:irrev}), which allow for fine-grained control over
  variable usage. Namely, linear resources occurring in non-WHNF case
  scrutinees can no longer be used in the alternative, while the
  pattern variables (or the case binder) must necessarily be used. } 
%
To the best of our
knowledge, this is the first type system \ROUNDTWO{to accept programs
which depend on non-strict evaluation to be understood as linear}
(Section~\ref{sec:main:linear-core}).

@@

Additional proofs and definitions are given in \ROUNDTWO{the extended
  version of the paper~\cite{techreport}.}
%Appendices~\ref{app:proofs},~\ref{app:optimisations} and~\ref{app:opsem}.

@@

Linear type systems guarantee that linear resources are consumed exactly once.
\ROUNDTWO{Thus}, whether a program is \ROUNDTWO{accepted as linear} intrinsically depends on
the definition of consuming a resource.
%
\ROUNDTWO{Traditionally}, consuming a resource \ROUNDTWO{has been} equated with its 
syntactic occurrence along a control-flow path in the program. However, as this
section \ROUNDTWO{aims to} make clear, \ROUNDTWO{in a non-strict context, a
syntactic definition of linearity is overly conservative}.
%
For instance, in the following Haskell-like program, \lstinline{f} is a linear
function (a function that must use its argument
exactly once), but there are two syntactic
occurrences of \lstinline{handle}: one is in a let-bound computation that
closes this handle, the other \ROUNDTWO{is in the then branch of a
  conditional}. If the program is 
evaluated lazily, the handle will, \ROUNDTWO{dynamically, either be
  used exactly once (i.e,~closed) in
the else branch, or be used exactly once (i.e.,~returned) in the then branch}, since the 
\ROUNDTWO{let-bound variable is not needed in that branch and
so the expression which closes the handle is never evaluated}.
%
\begin{lstlisting}
f : Handle *' $\lolli$ '* Handle
f handle =
    let closed = close handle in
    if <cond>
        then handle
        else closed
\end{lstlisting}
%
% Intuitively, a computation that depends on a linear resource to produce a
% result consumes that resource iff the result is effectively computed; in
% contrast, a computation that depends on a linear resource, but is never run,
% will not consume that resource.
%
% Even though there are two occurrences of \lstinline{handle} in the program, and
% it looks as though the \lstinline{handle} may be closed before returned, this function may
% actually only \emph{consume} used once -- it depends on the evaluation strategy of the
% language -- the \lstinline{handle} is closed iff the let bound computation is
% evaluated.
%
%
% Intuitively, \emph{consuming} a resource is not necessarily synonymous with its
% syntactic occurrence. In the example, \emph{non-strict} evaluation means the
% handle is never closed by the function.
%
%
% Depending on the evaluation strategy of the language, a computation that closes
% a handle might not be evaluated, and thus is not consumed.
% With \emph{eager} evaluation semantics,
% the let
%   bound expression \lstinline{close handle} is eagerly evaluated, and the \lstinline{handle} will be closed before being
% returned. It is clear that a linear type system should not accept such a
% program since the linear resource \lstinline{handle} is duplicated -- it is used in a
% computation that closes it, while still being made available to the caller of
% the function;
% %
% on the other hand, with \emph{call-by-need},
%
% Reconsider the \lstinline{handle} example program under
% \emph{call-by-need} evaluation semantics:
% The let bound expression is only evaluated if the binding \lstinline{x} is
% needed. The function returns the \lstinline{handle} right away, and forgets the
% let bound \lstinline{x} as its scope is exited, such that \lstinline{x} will
% never be needed -- consequently, the \lstinline{handle} is never closed by the
% function...
%
%
%

@@

\subsubsection{Branching on WHNF-ness}\label{sec:onwhnf}

@@

\subsubsection{Proof irrelevant resources}\label{sec:irrev}
